\bf Model Visualization Techniques for a Social Network Model

### 2018-04-03 paper-child.Rnw
edits

\section{Introduction}
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks.
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
\st{%XXX - okay, here is what i have done for that:
Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations. In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models. These models, however, are only for single instance networks. If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem. If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes. When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account.
Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be. A social network does not form spontaneously: it evolves over time. Ties are formed and dissolved, and new actors join the social structure. Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.

\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network. Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network. In the ``real" world we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this,
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.

>>>>

\section{Introduction}
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks.
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
%XXX - okay, here is what i have done for that:
Social networks, such as collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations have been studied for decades.  
%beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of . 
In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. %There are many kinds of social networks, but there are not as many statistical models for social network data. 
Some network models that have been applied to social networks include the exponential random graph model (ERGM) and latent space models. These models, however, are only for single instance networks. If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem. If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they are not able to model network change as time passes. When studying \textit{dynamic networks}, networks observed at many time points, we need a model that allows for variation in the network over time. 
Models for dynamic social networks have a great deal of potential because of how realistic they can be. A social network does not form spontaneously: it \textit{evolves} over time. Ties are formed and dissolved, and new actors join the social structure. Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.

\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network. Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network: we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.

%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit models, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this,
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user. In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes. %, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
